---
title: "04_cell_communication"
format: html
---


## Setup output folders

```{r}
#| label: setup-folders
#| include: false

# Automatically get today's date
today_date <- format(Sys.Date(), "%Y-%m-%d")

# Create directory path
output_dir <- file.path("results", "04_cell_comunication_outputs", today_date)

# Create folder if it doesn't exist
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

# Show where output files will go
output_dir
```

# Load packages

```{r}
#| label: load-packages
#| include: false
#| warning: false
#| message: false
#| error: false

library(CellChat)
library(Seurat)
library(dplyr)
library(ggplot2)
```

Perform cell chat communication analysis on the Banksy dataset using the cell type annotations and deconvolution results obtained from the previous step. This analysis will help us understand the cell-cell communication patterns in the spatial transcriptomics data.

# Load Data

Load data preprocessed with `03_cell_type_deconvolution.qmd` script. This data contains the cell type annotations for each spot in the Banksy dataset and deconvolution using scRNAseq from lymph node and PBMC from humans.

```{r}
object <- readRDS(file ="results/03_cell_type_deconvolution_outputs/2025-07-30/Banksy_merged_annotation.rds")
```

# Extract data to CellChat format


```{r}

# keep only high confidence deconvoluted spots

# Subset to keep only first_type and remove rejects/NA
object_subset <- subset(object, subset = !(spot_class %in% c("reject")) & !is.na(spot_class))


# Extract normalized data and metadata
data.input <- GetAssayData(object_subset, assay = "BANKSY", slot = "data")
coords <- GetTissueCoordinates(object_subset)
meta <- data.frame(first_type = object_subset$first_type)
rownames(meta) <- colnames(object_subset)
meta$first_type <- droplevels(as.factor(meta$first_type))
coords <- coords[colnames(data.input), ]  # Ensure same order
coords <- as.matrix(coords)
coords <- coords[,c(1,2)]

spatial.factors <- data.frame(
  ratio = 0.3,  # controls decay steepness (tune if needed)
  tol = 30      # ~15 µm in pixels, neighborhood radius for interactions
)

# Create CellChat object
DefaultAssay(object_subset) <- "BANKSY"
cellchat <- createCellChat(object = data.input,
                           meta = meta,
                           group.by = "first_type",
                           datatype = "spatial",
                           coordinates = coords,
                           spatial.factors = spatial.factors)

# Set the ligand-receptor database
CellChatDB <- CellChatDB.human 
cellchat@DB <- CellChatDB

```

# Preprocess data

```{r}
cellchat <- subsetData(cellchat)                        
cellchat <- identifyOverExpressedGenes(cellchat)
cellchat <- identifyOverExpressedInteractions(cellchat)
cellchat <- computeCommunProb(cellchat, raw.use = TRUE)
cellchat <- filterCommunication(cellchat, min.cells = 1)
cellchat <- computeCommunProbPathway(cellchat)
cellchat <- aggregateNet(cellchat)

```
```{r}
saveRDS(cellchat, file = file.path(output_dir, "cellchat_object.rds"))
```


# Save CellChat object

Save computed cell chat to avoid rerunninng

```{r}
saveRDS(cellchat, file = file.path(output_dir, "cellchat_object.rds"))
```

# Visualize calculated communication probabilities

## Visualize communication network

```{r}
groupSize <- as.numeric(table(cellchat@idents))

netVisual_circle(cellchat@net$count, vertex.weight = groupSize, 
                 weight.scale = TRUE, label.edge = FALSE,
                 title.name = "Number of interactions")

netVisual_circle(cellchat@net$weight, vertex.weight = groupSize, 
                 weight.scale = TRUE, label.edge = FALSE,
                 title.name = "Interaction strength")

```

## Visualize individual pathway (e.g., MIF)

```{r}
netVisual_aggregate(cellchat, signaling = "MIF", layout = "circle")
```


# Session Info
```{r}
sessionInfo()
```

